\textbf{Empirical Bayes shrinkage, and denoising of Poisson and heteroskedastic Gaussian signals}

### 2016-03-03 paper/supplementary/supplementary.tex
started writing supplementary materials

\section{Abstract}

\section{Introduction}

Shrinkage and sparsity play a key role in many areas of modern statistics, including, for example, high-dimensional regression \cite{Tibshirani1996Regression}, covariance or precision matrix estimation \cite{Bickel2008Covariance}, multiple testing \cite{efron:xx} and signal denoising \cite{Donoho1994Ideal, donoho95}. One attractive way to achieve shrinkage and sparsity is via Bayesian or Empirical Bayes (EB) methods (e.g.\cite{efron?, Johnstone2005Empirical, Clyde2000Flexible,george.mccullogh,bayesian-covariance-estimation}).  
However, these methods are usually perceived to require context-specific implementations, and this overhead can limit their use in practice. Here we consider a flexible EB approach to shrinkage, which we call {\it adaptive shrinkage}, whose goal is to provide a generic shrinkage method that could be useful for a range of applications. 
We show how this single shrinkage method can produce effective results for several denoising problems, including smoothing Gaussian means in the presence of heteroskedastic variance, smoothing of Gaussian variances, and smoothing Poisson means. These are all settings that are relatively underserved by existing EB implementations, and indeed  we are unaware of any existing EB implementation for smoothing either the mean or the variance in the heteroskedastic Gaussian case.
Consistent with previous studies (\cite{Antoniadis2001Wavelet}, \cite{Besbeas2004Comparative}) we find these EB methods to be more accurate than commonly-used thresholding rules, and, in the Poisson case, competitive with a purpose-built EB method.

Methods for smoothing the variance of Gaussian data, or even for smoothing the mean in the presence of heteroskedastic errors, are uncommon, particularly compared with the well-studied homoskedastic case. Previous non-wavelet-based work includes \cite{Fan1998Efficient}, who estimated the variance by smoothing the squared residuals using local polynomial smoothing, \cite{Brown2007Variance}, who employed difference-based kernel estimators, and \cite{Menictas2015Variational}, who developed a Bayesian mean field variational methodology for both mean and variance estimation. The only wavelet-based method for variance estimation we have found is \cite{Cai2008Adaptive}, which applied a wavelet thresholding approach to first order differences. In no case could we find publicly-available software implementations of these methods (but we did receive
code by email from \cite{Menictas2015Variational}, which we use in our comparisons).

For Poisson data, variance stabilizing techniques together with normal approximations have been proposed by \cite{D93} and \cite{Fryzlewicz2004HaarFisz}, by exploiting the mean-variance relationship of a Poisson distribution. \cite{Kolaczyk1996NonParametric} and \cite{kolaczyk1999wavelet} derived thresholds achieving optimal asymptotic properties in the context of wavelet transformations, similar to the thresholds in the i.i.d. Gaussian case.  However, \cite{Fryzlewicz2004HaarFisz} is computationally inefficient due to the presence of external cycle-spinning, and threshold-based methods may not result in satisfactory finite sample performance. Furthermore, all of these methods are sensitive to the choice of the primary resolution level as subsequent simulations will show. Multiscale analysis using recursive dyadic partitions within a Bayesian framework was later developed by \cite{Kolaczyk1999Bayesian} to make use of a particular form of likelihood factorization, but a relatively inflexible conjugate prior was chosen.

%Methods used to achieve shrinkage and sparsity include simple thresholding rules
%(e.g.~\cite{TI-thresh, SURE, others?,thresholding-outside-of-wavelets?}), penalized likelihood approaches (e.g.~\cite{lasso,elasticnet,glasso}), and 
%Bayesian or Empirical Bayesian (EB) methods (e.g.\cite{efron?, johnstone.silverman.xx,clyde.george.jrssb,george.mccullogh,bayesian-covariance-estimation}.
%While thresholding rules and penalized likelihood methods can possess good asymptotic properties in theory, empirical
%comparisons have found that Bayesian and EB methods can perform better in practice \cite{antoniadis:2001,guan.stephens.largescaleregression}.
%However, Bayesian and EB methods are usually perceived to require context-specific implementations, and this
%overhead can limit their application.
%Our goal here is to outline a strategy that can reduce this implementational overhead, and apply it to produce EB shrinkage methods for signal denoising problems.
%DO we want to highlight ash or the generic strategy? Or both? Current text highlights mostly the strategy,
%leaving ash as just a component of it.
%Could add "Indeed one of our goals is to showcase the flexibility of this EB approach to shrinkage" as second
%sentence of next paragraph?

%Among their most prominent
%applications has been signal denoising, which can be achieved by inducing sparsity or shrinkage in a transformed (e.g. wavelet)
%domain \cite{donoho.johnston.xx}. This approach to signal denoising has strong theoretical support, and is often accurate and computationally convenient, so it has
%become widely used. Consequently many approaches have been suggested for achieving the desired shrinkage, including both
%simple thresholding rules \cite{TI-thresh, SURE, others?} and Bayes or Empirical Bayes (EB) methods \cite{johnstone.silverman.xx,clyde.george.jrssb}.
%Although classical thresholding rules have been shown to possess good asymptotic properties in theory, EB methods often
%perform better in practice \cite{antoniadis:2001}.

\def\ash{{\tt ash}}
\def\bhat{\hat{\beta}}
\def\shat{\hat{s}}

The {\it adaptive shrinkage} (\ash) method is described in detail in a companion paper, \cite{stephens:2015}, which applies \ash~to estimate false discovery rates (FDR) in multiple testing settings. Here we apply the same method to the signal denoising problems described above.  In brief, \ash~aims to provide EB shrinkage estimates of quantities $\beta_1,\dots,\beta_J$, given observed estimates of these quantities $\bhat_1,\dots,\bhat_J$, and corresponding standard errors $\shat_1,\dots,\shat_J$. In common with most EB methods \ash~assumes that the $\beta_j$ come from some common underlying distribution, $g$, which is estimated from all the data (as $\hat{g}$ say), and uses the posterior mean,  $E(\beta_j | \bhat_j, \shat_j, \hat{g})$ as a shrinkage estimate of $\beta_j$. Key features that make \ash~attractive as a flexible and generic approach to shrinkage include: i) its flexible semi-parametric modelling of $g$, under the constraint that it be uni-modal at 0 (and, optionally, symmetric); ii) its incorporation of the precision of each measurement through the standard errors $\shat_j$, so that the amount of shrinkage adapts to the precision of each measurement;  iii) its computational simplicity and speed: e.g.~our {\tt R} implementation typically takes a fraction of a second for $J=1000$; iv) its computational stability, with, for example, minimal problems due to convergence to local optima \cite{stephens:2015}.

% treating the observations as conditionally independent from a normal model, $\bhat_j | \beta_j, \shat_j \sim N(\beta_j, \shat_j)$. Given the estimate $\hat{g}$ for $g$, \ash~then computes the posterior distiributions $\beta_j | \bhat_j, \shat_j, \hat{g}$,vand provides the posterior mean of $\beta_j$ as a shrinkage estimate for $\beta_j$.

%The name ``adaptive shrinkage" comes from the fact that shrinkage is adaptive to both the overall signal in the data (through estimation of $g$), and to the precision of each observation (through incorporation of $\shat_j$), with imprecise observations being shrunk most strongly.

The \ash~method makes assumptions, particularly conditional independence of observations, and normality of likelihood $p(\bhat_j | \beta_j, \shat_j)$, that will often be violated in practice,
and indeed are violated in our applications here. However, as we demonstrate later, it can nonetheless produce good shrinkage estimates. 
Essentially, we treat
%highlighting the potential of treating ??
\ash~as a generic or ``black box" shrinkage procedure, which inputs
estimates and their corresponding standard errors and outputs shrinkage estimates, without worrying too much about 
the details. Of course, one must be careful about how far one takes this.  And in more complex applications 
(e.g. large-scale regression and covariance estimation) it is unclear whether this strategy can be usefully applied. However,
we believe our results here for signal denoising, together with those in \cite{stephens:15} for FDR estimation, illustrate \ash's potential and flexibility. 
%maybe move to discussion

%In this paper we use \ash~to tackle several signal denoising problems,  including Poisson smoothing,
%and smoothing both the mean and variance of heteroskedastic Gaussian data. 
Shrinkage methods are widely-used in signal denoising applications, because signal denoising can be accurately and conveniently achieved by shrinkage in a transformed (e.g.~wavelet) domain \cite{Donoho1994Ideal}. 
Commonly-used shrinkage methods include both
simple thresholding rules \cite{Coifman1995Translationinvariant, Donoho1994Ideal, donoho95} and EB methods \cite{Johnstone2005Empirical, Clyde2000Flexible}.
Being an EB method, \ash~has much in common with these previous EB methods, but generalizes them in two ways: first, 
\ash~allows more flexibility in the underlying distribution $g$ than the Laplace or spike-and-slab
distributions used in \cite{Johnstone2005Empirical, Clyde2000Flexible}; second, \ash~allows for variations in precision in the transformed observations (e.g. wavelet coefficients). The latter property is particularly important for
the Poisson and heteroskedastic Gaussian settings we consider here.

Software implementations of our methods are available in the R packages {\tt ashr} (Adaptive SHrinkage in R)
and {\tt smash} (SMoothing by Adaptive SHrinkage), available from 
\url{http://www.github.com/stephens999/ashr} and \url{http://www.github.com/stephenslab/smash} respectively.

%Although classical thresholding rules have been shown to possess good asymptotic properties in theory, EB methods often
%perform better in practice \cite{antoniadis:2001}.

%Our goals in this paper are therefore two-fold. FIrst, to showcase the potential for the adaptive shrinkage method to provide a simple and effective
% solution to shrinkage estimation in important practical problems; and, second, to provide software implementations of this 
% idea for smoothing Poisson and heteroskedastic Gaussian signals - settings that are underserved by existing implementations.

%In recent years, various shrinkage and thresholding procedures have become ever more popular in many applications where sparsity is desired. Here ``sparsity'' is loosely defined in that it also includes the case where many terms are near-zero, as opposed to exactly zero. The most prominent usage of such procedures appear in multiresolution denoising analyses such as wavelets, where sparsity in a transformed domain often corresponds to smoothness in the data space [references]. Other examples include estimation of covariance and precision matrices (though thresholding is used more often here since zero terms are actually desired) [references], accounting for different measurement precisions in FDR control for multiple testing [references], or more generally optimizing penalized likelihood problems such as LASSO [references]. In this paper we will look at a flexible and adaptive shrinkage method named ASH, and illustrate its value in two key areas of univariate wavelet denoising.

%Before delving deeper into the two denoising tasks, we first present a brief overview of wavelet estimators. These estimators have become extremely popular in nonparametric regression problems following the seminal paper by Donoho \& Johnstone (1994). Wavelet methods are locally adaptive in that they achieve the optimal minimax rate over a wide variety of functions, yet are also computationally faster than many other adaptive methods such as variable-bandwidth kernel methods [antoniadis 2001]. While classical thresholding estimators have been shown to possess optimal asymptotic properties, an extensive simulation study by Antoniadis et al. (2001) has demonstrated that various Bayesian shrinkage methods are usually more adaptive than thresholding methods in finite samples, outperforming the latter in terms of mean squared error (MSE). However, many Bayesian methods require specific distributional assumptions, and are not easily extensible to more complex problems, two of which we subsequently describe. On the other hand, our method simplies the task by making approximations, while still possessing good finite sample performance and computational efficiency. These approximations also allow us to use ASH as the common shrinkage procedure in both the problems.

%Both the aforementioned problems are harder to deal with than the classical problem with i.i.d. Gaussian errors, and usually requires extension of the ideas from the latter task. Before proceeding to describe the two problems in detail, we will first describe briefly the portion of ASH that performs shrinkage.

%The flexibility and computational efficiency of ASH is immediately obvious, which motivates us to extend this method to the more complex wavelet settings described here. As an Empirical Bayes procedure, ASH borrows information across all available data and is extremely data adaptive. At the same time, the key advantage of ASH over other Empirical Bayes procedures such as EbayesThresh [Johnstone \& Silverman (2005)] and the commonly used ``spike-and-slab'' prior (is there a reference for this?) is that variable variances are allowed as input, a feature which latter methods lack. This is also the primary reason we can use ASH as the only shrinkage procedure in both the problems described above. Furthermore, Gaussian mixtures can effectively approximate any unimodal distribution, allowing for an extremely flexible prior. In tackling these two specific problems, we can speed up the entire signal denoising problem via likelihood approximations (discussed in the next section), allowing our method to be computationally efficient without compromising accuracy. Typical wavelet methods require the specification of a primary resolution level as a ``tuning'' parameter, which could substantially influence the accuracy of the method. However, applying ASH to every resolution in the wavelet transformationwe spares us the need for such a parameter. For an in-depth understanding of the original motivations and applications of ASH, see Stephens (??). With the main shrinkage method accounted for, we will proceed to explore the two aforementioned problems in detail. Note that one only needs to supply a vector of estimates $\hat{\Gb}$ and their standard errors se$(\hat{\Gb})$ to ASH to obtain posterior estimates; hence, we will focus on obtaining these estimates and standard errors when describing our methods.

\section{Methods}

\def\bY{\bm{Y}}
\def\bmu{\bm{\mu}}
\def\bp{\bm{p}}
\def\be{\bm{\epsilon}}
\def\tbY{\tilde\bY}
\def\tbe{\tilde\be}
\def\tbmu{\tilde\bmu}
\def\tmu{\tilde\mu}
\def\tY{\tilde{Y}}
\def\s{\sigma}
\def\|{\,|\,}

\subsection{Adaptive shrinkage}

Here we briefly outline the adaptive shrinkage method; see \cite{stephens:2015} for full details.
Adaptive shrinkage (\ash) is an EB method
for estimating quantities $\beta=(\beta_1,\dots,\beta_n)$ from noisy estimates 
$\bhat=(\bhat_1,\dots,\bhat_n)$ and their corresponding standard errors $\shat = (\shat_1,\dots,\shat_n)$. 
In its simpest form it assumes the hierarchical model
\begin{align}
\label{eq:ash_model_1}\beta_j \| \shat_j & \sim g \\
\bhat_j \| \beta_j, \shat_j & \sim N(\beta_j, \shat_j^2),
\end{align}
where the distribution $g$ is constrained to be unimodal and symmetric. This constraint on $g$ can
be flexibly achieved using 
a mixture of zero-centered normal distributions

\begin{equation}
g(\cdot)=\sum_{k=0}^K \pi_0 N(\cdot;0,\sigma_k^2),
\end{equation}
where the mixture weights $\pi_0,\dots,\pi_K$ are non-negative and sum to 1, and $N(\cdot; \mu,\sigma^2)$ denotes
the density of a normal distribution with mean $\mu$ and variance $\sigma^2$.
A key idea, which substantially simplies inference, is to take
$\sigma_0,\dots,\sigma_K$ to be a fixed grid of values. ranging from very small (e.g.~$\sigma_0=0$,
in which case $g$ includes a point mass at 0) to very large. Estimating $g$ then boils down to estimating
the mixture weights $\pi$, which is done by maximum likelihood using a simple EM algorithm.
Given an estimate $\hat{g}$ for $g$, the conditional distributions $p(\beta_j \| \bhat, \shat, \hat{g})$ are analytically
tractable, and the posterior mean $E(\beta_j \| \bhat, \shat, \hat{g})$ provides a shrinkage point estimate for $\beta_j$.

\cite{stephens:2015} also introduces various embellishments that are implemented in the {\tt ashr}
package, including generalizing the normal likelihood to a $t$ likelihood, and dropping the symmetric
constraint on $g$ by replacing the mixture
of normals with a more flexible (though less smooth) mixture of uniforms. However, we do not use these embellishments here.

%\cite{stephens:2015} demonstrates the potential for this adaptive shrinkage approach to  estimate false discovery rates and related quantities in settings where the primary goal is to identify the $\beta_j$ that differ ``significantly" from 0. Our work here compliments this, by demonstrating  the potential for this approach to provide the adaptive shrinkage estimates that its name promises.

\subsection{Signal Denoising}

A common generic signal denoising problem, sometimes known as ``non-parametric regression", involves
estimating a ``spatially-structured" mean $\bmu=(\mu_1,\dots,\mu_T)$, from corresponding
noisy observations $\bY = (Y_1,\dots,Y_T)$, where $t=1,\dots,T$ indexes location in a one-dimensional space,
such as time, or, as in our example later, location along the genome. 
 By ``spatially-structured" we mean that $\mu_t$ will often be similar to $\mu_t'$ for small $|t-t'|$, though we do not rule
out occasional abrupt changes in $\mu$.  For convenience we assume that $T=2^J$ for some integer $J$, as is common when using wavelet methods.

The most studied denoising problem is the homoskedastic Gaussian case; that is, 
the data $Y_t$ have Gaussian noise and constant variance.
Here we consider the more general case of Gaussian data with
spatially-structured mean {\it and} spatially-structured (non-constant) variance. 
In some settings the changes in variance may themselves be of interest, and our methods provide explicit estimates for the variance as well as the mean. In addition we consider Poisson data (where the variance depends on the mean, so 
a spatially-structure mean implies spatially-structure variance).
To build up to these more interesting cases we start with the simplest setting: 
Gaussian noise with known variance.

\subsubsection{Gaussian noise; estimate mean (known variance)}

Suppose the $Y_t$ are independent noisy observations of the $\mu_t$, with Gaussian noise of known variance $\s^2_t$. That is,
\begin{equation}\label{eq:1d gaussian model}
\bY =\bmu+\be
\end{equation}
where $\be \sim N(0,D)$ with $D$ the diagonal matrix with diagonal entries $(\s_1^2,\dots,\s_T^2)$.

%To motivate our method, we first describe wavelet shrinkage from a Bayesian perspective.
Wavelet denoising involves first applying a discrete wavelet transform to the data $\bY$. This
involves pre-multiplying $\bY$ by an orthogonal $n\times n$ matrix $W$ that depends on the orthonormal wavelet basis chosen. Pre-multiplying (\ref{eq:1d gaussian model}) by $W$ yields
\begin{equation} 
W\bY=W\bmu + W\be 
\end{equation}
which, using $\,\tilde{}\,$ to denote the wavelet transform, we write
\begin{equation} \label{eq:wc}
\tbY=\tbmu + \tbe,
\end{equation}
where $\tbe \sim N(0,W D W')$.

A key feature of the wavelet transform is that if $\bmu$ is spatially structured 
then many elements of $\tbmu=W\bmu$ will tend to be close to zero, and vice versa.
In other words, smoothing in the data domain corresponds to shrinking in the wavelet domain.
Thus, one can obtain a spatially structured estimate of
mean $\bmu$ by first using shrinkage methods to estimate $\tbmu$ and then
reversing the wavelet transform.

%This process is often referred to as ``wavelet shrinkage estimation", because the estimates $\tmu_j$ are obtained from $\tY_j$ by ``shrinking" them towards zero.

Here we use the adaptive shrinkage method to obtain shrinkage estimates for $\tbmu$.
Focussing on the marginals of (\ref{eq:wc}), we have
\begin{equation} \label{eq:normlik}
\tY_j | \tmu_j, s^2_j \sim N(\tmu_j, \omega^2_j).
\end{equation}
%Note that the elements of $\tY$ are not independent unless the variances $\sigma^2_t$ are all equal. 
where
\begin{equation}
\omega^2_j:=\sum_{t=1}^T \s^2_t W_{jt}^2,
\end{equation}
are the diagonal elements of $WDW'$.
Thus, applying ash with $\bhat_j = \tY_j$ and $\shat_j = \omega_j$ yields shrinkage estimates for the $\tmu_j$.
(In practice it is important to group the wavelet-transformed observations $j$ 
by their resolution level before shrinking; see note below.)
Of course, by focusing on the marginals we are ignoring any correlations among the $\tY_j$, and this
is the primary simplification here. (We are not alone in making this simplification;
see also \cite{Silverman1999Wavelets} for example.)

The above outlines the basic strategy, but there are some important
 additional implementational details:
\begin{itemize}
\item Rather than use a single wavelet transform, we use the ``non-decimated" wavelet transform, which treats the observations as coming
from a circle, rather than a line, and averages results over all $T$ possible rotations of the data. Although not always necessary, this is a standard trick to reduce artifacts that can occur near discontinuities in the underlying signal (eg. \cite{Coifman1995Translationinvariant}). 
%COMMENT: WHat about reflection before doing this? ans: all the test fns are periodic
%DO OTHER EB APPROACHES CONSIDER THE NDWT? yes ebayesthresh software discusses this option
\item The non-decimated wavelet transform yields $T$ wavelet coefficients (transformed values of $\bY$) 
at each of $J=\log_2(T)$ resolution levels.
We apply ash separately to the wavelet coefficients at each resolution level, so that
a different distribution $g$ for the $(\tmu_j)$ is estimated for each resolution. This is the usual way that
EB approaches are applied in this context (e.g.~\cite{Johnstone2005Empirical}) and indeed is crucial because
the underlying distribution $g$  will vary with resolution (because smoothness of $\bmu$ will vary with resolution).  
\item Although we have presented the wavelet transform as a matrix multiplication, which is an $o(T^2)$ operation, in practice both the wavelet transform and the inverse transform are implemented using
efficient algorithms \cite{Beylkin1992Ontherepresentation,Coifman1995Translationinvariant}, implemented in the {\tt wavethresh} package \cite{Nason_wavethresh}, taking only $O(T\log T)$ operations. 
\end{itemize}
%\item Ash yields a posterior distribution for each $\tmu_t$. A simple approach would
%be to use the posterior mean for $\tmu_t$ and then invert the wavelet transform; however we instead use the delta method?

%Since $\mathbb{V}(Y)=\s^2I$,
%\begin{eqnarray}\label{eq:waveletcoef}
%\bm{d}\sim N_n(\bm{\Ga},\s^2I)
%\end{eqnarray}
%where $\bm{\Ga}=W\mathbb{E}(\bm{Y})$. This implies that the likelihood for $\bm{\Ga}$ factorizes into a product of likelihoods for $\Ga_t$, $i=1,...,n$. As such, a natural and computationally convenient approach is to set independent priors on $\Ga_t$ as a form of ``shrinkage''. Here we apply the priors from ASH to each resolution level separately, resulting in
%%Although there are many choices for the prior, we use the Gaussian mixture prior from ASH in our method, given by
%\begin{eqnarray}\label{eq:ashprior}
%\Ga_{jk}=\sum_{l=1}^m \pi_l^{(j)} N(0,(s_l^{(j)})^2)
%\end{eqnarray}
%with $\sum_l \pi_l^{(j)}=1$, for $j=1,...,J$. $\pi_t^{(j)}$ and $(s_t^{(j)})^2$ are hyperparameters that are shared between coefficients in the same resolution level, and $m$ is the number of mixture components. In ASH, $m$ is usually chosen to reasonably approximate any unimodal distribution, while maintaining an acceptable computational speed. Note that, for notational convenience, we have switched to a double index following standard wavelet convention for indices: here $j=1,...,J$ is the resolution level, and $k=0,...,2^j-1$ is the location within each resolution level $j$. By applying the inverse wavelet transform to the posterior mean $\hat{\bm{\Ga}}$ that ASH produces, one can obtain the posterior mean of $\bm{\mu}$, which serves as an estimate that minimizes the MSE. One could also construct credible bands using the posterior variances.
%
%While many existing methods perform well in the presence of i.i.d. (Gaussian) errors, heteroskedastic (but still independent) errors present a different challenge. One key obstacle when dealing with heteroskedastic errors is that the likelihood for $\bm{\Ga}$ does not necessarily factorize. Given the true variances and hence the likelihood, one could compute the full posterior distribution using some specified prior. However, this would be computationally cumbersome, and may be infeasible when extended to multiple signals or higher dimensions. In this case, a suitable prior might also be difficult to find. As such, one key aspect of our approach is to treat the wavelet coefficients as if they were independent, so that the true likelihood is approximated by a composite likelihood (see Silverman (1999) for an example where this is done).

%To describe our approach in more detail, first assume that the true variance function is known and given by $\mathbb{V}(\Ge_t)=\s_t$, and that the true mean function is given by $\bm{\mu}=(\mu_1,...,\mu_t)$, where $T=2^J$ for some positive integer $J$. Hence
%\begin{eqnarray}
%\bm{Y}\sim N_n(\bm{\mu},\Sigma)
%\end{eqnarray}
%where $\Sigma=\textrm{diag}(\s_1,...,\s_T)$ is a diagonal matrix. For the rest of this subsection, we will use the non-decimated wavelet transform (NDWT) instead of the standard DWT to achieve translation invariance. This in turn reduces the presence of artifacts, which often occur near discontinuities in the underlying signal (see eg. Coifman \& Donoho (1995)). Then,
%\begin{eqnarray}
%\bm{d}\sim N_n(\bm{\Ga},\tilde{\Sigma})
%\end{eqnarray}
%where $\bm{d}$ is the vector of detail coefficients, $\bm{\Ga}=W\bm{\mu}$, and $\tilde{\Sigma}=W\Sigma W^T$, where $W$ is the matrix associated with the NDWT for a given wavelet basis. By treating the likelihood for $\bm{\Ga}$ as if it were independent, it can be written as follows(using the double index mentioned above):
%\begin{eqnarray}\label{eq:likelihood}
%L(\bm{\Ga}|\bm{d})=\prod_{j=0}^J\prod_{k=0}^{T-1}P(d_{jk}|\Ga_{jk})
%\end{eqnarray}
%where $P(d_{jk}|\Ga_{jk})=\phi(d_{jk};\Ga_{jk},\tilde{\Sigma}_{(jk,jk)})$. Note that there are $n$ coefficients at each resolution level instead of $2^j$ for resolution $j$ due to the NDWT. Here $\phi$ denotes the Gaussian density function. Since $\Sigma$ is diagonal, it is easy to see that $\tilde{\Sigma}_{(jk,jk)}=\sum_{t=1}^T \Sigma_{ti}_{tW_{jk,i}^2$. In our method, we use ASH to assign independent priors to $\Ga_{jk}$ as with \eqref{eq:ashprior}. %ASH also allows us to estimate the $\pi_t$'s via a maximum likelihood approach and fix the $\s_t$'s by using a grid of values. This particular choice of prior specification is extremely flexible, being able to approximate any symmetrical zero-centered distribution, as well as allowing for a relatively easy estimation of the hyperparameters through empirical Bayes. See ASH for more details regarding this choice of prior (?)
%While ASH produces shrunken estimates of $\Ga_{jk}$ for $j,k>0$, we estimate $\Ga_{00}$ using the corresponding scaling coefficient, which seems intuitively appealing. Finally, we can obtain an estimate of $\mu$ by using the average basis inverse, which is essentially an average of the inverse wavelet transforms for every shift of the data (see Coifman \& Donoho (1995)) since we are using the NDWT. Although the method described here uses a matrix formulation for easier conceptual understanding, the actual NDWT and inverse transform are done through Mallat's pyramid algorithm, taking only $n\log(n)$ time.

\subsubsection{Estimating spatially structured variance (known mean)}

Although we assumed variances to be known above, the problem of variance estimation is itself non-trivial. 
A common approach is to assume that the variance is constant ($\sigma_t=\sigma$), and that changes in the adjacent means $\mu_t-\mu_{t+1}$ are negligible, and to use an unbiased estimate for $\sigma$ such as
\begin{equation} \label{eq:constvar}
\hat{\sigma}^2 = (1/2T) \sum_{t=1}^{T} (Y_t-Y_{t-1})^2
\end{equation}
with $Y_0:=Y_T$.
Here we make the more flexible assumption that the variance function is spatially structured, and use wavelet shrinkage to estimate it. Our approach is similar in spirit to \cite{Delouille2004Smooth}, and also to \cite{Cai2008Adaptive} (although the latter estimates only the variance function).

Assume initially that the mean $\mu_t$ is known, and define
\begin{eqnarray}\label{eq:varobs1}
Z_t^2=(Y_t-\mu_t)^2_t
\end{eqnarray}
to be the ``observations'' for the unknown variance function. Note that $\mathbb{E}(Z_t^2)=\s_t^2$, and so
we have a mean estimation problem. To tackle this we use the mean estimation procedure above (effectively treating the wavelet-transformed values $WZ_t^2$ as Gaussian when really they are linear combinations of $\chi^2$ random variables). To apply this procedure we need the variance
of $Z_t^2$, $\mathbb{V}(Z_t^2)$, which is unknown.
Here we use $\frac{2}{3}Z_t^4$ as an 
unbiased estimator for $\mathbb{V}(Z_t^2)$. (This follows from the distributional result that, when $Z^2 \sim \s^2 \chi^2$, then $\mathbb{E}(Z^4)= 3\s^4$, and $\mathbb{V}(Z^2)=2\s^4$.)

Despite the approximations being made here, we have found this procedure to work well in practice in most cases,
perhaps with a tendancy to oversmooth quickly-varying variance functions. 
%Although this works well in most cases, there are variance functions for which the above procedure tends to overshrink the detail coefficients at the finer levels. This is likely because the distribution of the wavelet coefficients are extremely skewed, especially when the true coefficients are large (at coarser levels the distributions are much less skewed since we are dealing a linear combination of a large number of data points).

%MOVE THIS TO DISCUSSION?
%Of course, better likelihoods and priors could be used that reflect the skewness in the distributions of the ``observations'', but we found that a normal approximation works reasonably well (especially for smoother variance functions), and has the key advantage of being easy and fast to implement. The variance estimation process is then very similar to one for mean estimation as described above, with a few extra details. The full procedure is outlined in Appendix \ref{app:var estimation}.

\subsubsection{Estimating spatially-structured mean and variance}

Having specified procedures for estimating the means with variances known, and the variances with means known,
we iterate these procedures to deal with the (typical) case where both are unknown.
We initialize the algorithm by estimating the variance vector $\bm{\s}^2$ using
\begin{eqnarray}\label{eq:initial var est}
\hat{\s}_t^2=\frac{1}{2}\left((Y_t-Y_{t-1})^2+(Y_t-Y_{t+1})^2\right)
\end{eqnarray}
where $Y_0\equiv Y_n$ and $Y_{T+1}\equiv Y_1$ (equivalent to putting the observations on a circle). 
%This estimator has the property that it is approximately unbiased for the variance at the two corresponding points, provided that the mean and the variance function is smooth enough.
Then we iterate:
\begin{enumerate}
\item[1] Estimate $\bm{\mu}$ as if $\bm{\s}^2$ is known (with the value obtained from the previous step).
\item[2] Estimate $\bm{\s}^2$ as if ${\bm{\mu}}$ is known (with the value obtained by the previous step); return to 1.
\end{enumerate}
We cannot guarantee that this procedure will converge in general. In our simulations we found
that two iterations of steps 1-2 yielded accurate results (so the full procedure consists of initialize + Steps 1-2-1-2).

 \subsection{Smoothing Poisson data}

\def\Poi{\text{Poi}}
\def\Bin{\text{Bin}}

Now assume that each $Y_t$ has a Poisson distribution with mean $\mu_t$.
To do denoising here we apply adaptive shrinkage to the Poisson multiscale models
from \cite{Nowak2000Statistical} and \cite{Timmermann1999Multiscale}, which are an analogue of
wavelet methods for Poisson data.

To explain the idea, first recall the following elementary distributional result:
if $Y_1,Y_2$ are independent, with $Y_j \sim \Poi(\mu_j)$ then 
\begin{align}
Y_1 + Y_2 & \sim \Poi(\mu_1 + \mu_2) \\
Y_1 | (Y_1+Y_2) & \sim \Bin(Y_1+Y_2, \mu_1/(\mu_1+\mu_2)).
\end{align}

To extend this to $T=4$, introduce the notation $v_{i:j}$ to denote, for any vector $v$, 
the sum $\sum_{t=i}^j v_t$.  Then 
\begin{align} 
Y_{1:4} & \sim \Poi(\mu_{1:4}) \label{eqn:y14} \\
Y_{1:2} | Y_{1:4} & \sim \Bin(Y_{1:4}, \mu_{1:2}/\mu_{1:4})  \label{eqn:y12} \\
Y_1 | Y_{1:2} & \sim \Bin(Y_{1:2}, \mu_1/\mu_{1:2}) \label{eqn:y1} \\ 
Y_3 | Y_{3:4} & \sim \Bin(Y_{3:4}, \mu_3/\mu_{3:4}). \label{eqn:y3}
\end{align} 
Together these models are exactly equivalent to $Y_j \sim \Poi(\mu_j)$, and
they decompose the overall distribution $Y_1,\dots,Y_4$ into parts involving aspects of the data at 
increasing resolution: (\ref{eqn:y14}) represents the coarsest resolution (the sum of all data points), 
whereas (\ref{eqn:y1}) and (\ref{eqn:y3}) represent the finest resolution, with (\ref{eqn:y12}) in between. 
Further, this representation suggests a reparameterization, from $\bmu=(\mu_1,\mu_2,\mu_3,\mu_4)$
to $\mu_{1:4}$ plus the binomial parameters $\bp=(\mu_{1:2}/\mu_{1:4}, \mu_1/\mu_{1:2}, \mu_3/\mu_{3:4})$, where 
$p_1$ controls lower-resolution changes in the mean vector $\mu$ and $p_2,p_3$ control higher resolution changes.

This idea extends naturally to $T=2^J$ for any $J$, reparameterizing $\bmu$ into its sum $\mu_{1:T}$
and a vector $\bp$ of $T-1$ binomial probabilities that capture features of $\bmu$ at different resolutions.
This can be thought of as an analogue of the wavelet transform of $\bmu$ for Poisson data.

Note that, in this reparameterization, $p_j=0.5 \, \forall j$ corresponds to the case of a constant mean vector,
and values of $p_j$ far from 0.5 correspond to large changes in $\mu$ (at some scale).
Thus estimating a spatially-structured $\bmu$ can be achieved by shrinkage estimation of $\bp$, with shrinkage towards $p_j=0.5$. Both \cite{kolaczyk1999wavelet} and \cite{Timmermann1999Multiscale} use purpose-built Bayesian models to
achieve this shrinkage, by introducing a prior distribution on elements of $\bp$ that is
a mixture of a point mass at 0.5 (creating shrinkage toward 0.5) 
and a Beta distribution. Here we take a different approach, reparameterizing $\alpha_j = \log(p_j/(1-p_j))$,
and then using adaptive shrinkage to shrink $\alpha_j$ towards 0.

To apply adaptive shrinkage we need an estimate $\hat{\alpha}_{j}$ and corresponding standard error $\hat{s}_j$
for each $j$. This boils down to estimating a log-odds ratio, and its standard error, which is a well-studied problem
(e.g. \cite{Gart1967Bias}). The main challenge is in dealing satisfactorily with cases where the maximum likelihood estimator (MLE) for $\alpha_j$ is infinite. Our choice of estimates, based on results from \cite{Gart1967Bias}, are described in Appendix \ref{}.

>>>>

\section{Simulation study for Gaussian data}
As noted in the main text, we performed an extensive simulation study for Gaussian errors using a variety of test functions (7 mean functions and 5 variance functions, including constant variance), signal-to-noise ratios (SNR = 1 and 3) and sample sizes ($T=256,512,1024$). The results for $T=256,512$ are saved under ~/res_paper/res_gaus_256.RData and ~/res_paper/res_gaus_512.RData respectively, while results for $T=1024$ are presented in a DSC framework (https://github.com/stephens999/dscr), saved under ~/res_paper/res_gaus_dscr.RData. In particular, we include an RShiny app as an interactive way to present the results for $T=1024$, the code for which can be found in the dscr-smash repo (https://github.com/zrxing/dscr-smash), under graphs.Rmd. In particular, users can view the boxplots of MISEs for the methods and test functions they are interested in by checking the appropriate boxes. For more details refer to graphs.Rmd.

In addition to the performance of the various methods, Table \ref{} also provides additional information for each of the method that we used in the simulation study.

\section{Simulation study for Poisson data}
The results from the simulation study for Poisson noise are shown in the following tables. Each of the six tables presents results for the corresponding mean function, and includes results for all three (min,max) intensity levels ((0.01,3), (1/8,8), (1/128,128)).

In addition we explore the results from trying out different Gaussian de-noising procedures for the second step of the Haar-Fisz (HF) algorithm.
